---
title: "Introduction to Bioconductor-1Exercise"
author: "@sungml92"
date: "5/25/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# load packages
library(GenomicRanges)
library(AnnotationHub)
```

# Exercise 1

## CpG islands on autosome

In this exercise, we are going to apply our knowledge in AnnotationHub package. We are to obtain data on "CpG Island" in human genome.

```{r}
ahub = AnnotationHub()

# target human genome
ahub = subset(ahub,species=="Homo sapiens")

# query CpG islands
CpG = query(ahub,"CpG islands")
CpG$genome
CpGhg19 <- CpG[[1]] # get hg19
CpGshg19 <- keepStandardChromosomes(CpGhg19,pruning.mode="coarse") # remove unusual chromosomes

# subset hg19 CpG islands in autosome
autosomeCpG <- dropSeqlevels(CpGshg19,c("chrX","chrY","chrM"),pruning.mode="coarse") # subset automsome
autosomeCpG
```

There are 26641 CpG islands exists on the autosomes

When we continue and practice additional filter..

```{r}
# filter on chr4
chr4CpG <- keepSeqlevels(autosomeCpG,"chr4",pruning.mode="coarse")
chr4CpG
```

There are 1031 CpG islands on chr4.

## H3K4me3 on autosome

Now we will obtain data for H3K4me3 histone modification for the H1 cell line from Epigenomics Roadmap and subset to include modifications only on autosomes.

```{r}
H3K4me3 <- query(ahub,c("H3K4me3","EpigenomeRoadMap"))

# search for H1 cell line and narrow peak
H1cellH3K4me3 <- H3K4me3[["AH29884"]]
# remove unusual chromosomes
H1cellH3K4me3stand <- keepStandardChromosomes(H1cellH3K4me3,pruning.mode="coarse") 
# subset automsome
#autosomeH3K4me3 <- subset(H1cellH3K4me3stand, seqnames %in% c(paste("chr", 1:22, sep="")))
autosomeH3K4me3 <- dropSeqlevels(H1cellH3K4me3stand,c("chrX","chrY","chrM"),pruning.mode="coarse") # subset automsome
sum(width(autosomeH3K4me3))
```

There are 41135164 bases covering autosome

## H3K27me3 on autosome

Now we will obtain data for H3K27me3 histone modification for the H1 cell line from Epigenomics Roadmap and subset to include modifications only on autosomes.

```{r}
H3K27me3 <- query(ahub,c("H3K27me3","EpigenomeRoadMap"))
# search for H1 cell line and narrow peak
H1cellH3K27me3 <- H3K27me3[["AH29892"]]
# remove unusual chromosomes
H1cellH3K27me3stand <- keepStandardChromosomes(H1cellH3K27me3,pruning.mode="coarse") 
# subset automsome
#autosomeH3K27me3 <- subset(H1cellH3K27me3stand, seqnames %in% c(paste("chr", 1:22, sep="")))
autosomeH3K27me3 <- dropSeqlevels(H1cellH3K27me3stand,c("chrX","chrY","chrM"),pruning.mode="coarse") # subset automsome
mean(autosomeH3K27me3$signalValue)

```

Mean single value for H3K27me3 is 4.770728

## Bivalent regions

Bivalent regions are bound by both H3K4me3 and H3K27me3

```{r}
bivalent <- intersect(autosomeH3K4me3,autosomeH3K27me3)
seqnames(bivalent)
sum(width(reduce(bivalent)))
```

10289096 bases on autosomes are covered by bivalent modifications

## CpG + Bivalent

```{r}
CpGbivalent <- subsetByOverlaps(bivalent,autosomeCpG,ignore.strand=TRUE)
length(CpGbivalent)/length(bivalent)

sum(width(CpGbivalent))/sum(width(autosomeCpG))
```
53.83644% of the bivalent regions overlap one or more CpG islands
24.1688% part of CpG islands bases are also bivalent marked

```{r}
CpG_10K <- resize(autosomeCpG,width=20000+width(autosomeCpG),fix="center")
CpG10Kbivalent <- intersect(bivalent,CpG_10K)
sum(width(CpG10Kbivalent))
```

9782086 bases are bivalently marked within 10Kb of CpG islands

## Human genome and CpG island

```{r}
genome <- ahub[["AH5018"]]
sgenome <- keepStandardChromosomes(genome,pruning.mode="coarse") 
# subset automsome
autogenome <- dropSeqlevels(sgenome,c("chrX","chrY","chrM"),pruning.mode="coarse") # subset automsome

autogenome_size <- sum(as.numeric(seqlengths(autogenome)))
sum(width(autosomeCpG))/autogenome_size
```

0.7047481 human genome were contained in CpG island

## Odds-ratio

```{r}
inOut = matrix(0,ncol=2,nrow=2)
colnames(inOut) = c("in","out") 
rownames(inOut) = c("in","out")

inOut[1,1] = sum(width(intersect(autosomeCpG,bivalent,ignore.strand=TRUE)))
inOut[1,2] = sum(width(setdiff(autosomeCpG,bivalent,ignore.strand=TRUE)))
inOut[2,1] = sum(width(setdiff(bivalent,autosomeCpG,ignore.strand=TRUE)))
# Last we need to fill out number of human genomes that are not part of this
inOut[2,2] = autogenome_size - sum(inOut)

oddsRatio = inOut[1,1] * inOut[2,2] / (inOut[2,1] * inOut[1,2])
oddsRatio
```

odds-ratio for the overlap of bivalent marks with CpG islands is 169.0962